Wigglyhedra

### 2025-01-02 MZ/answerReferee.tex

We are grateful to the referee for their thorough reading and their constructive suggestions on our manuscript \emph{Wigglyhedra}. We would like to submit a revised version of this manuscript to the \emph{Mathematische Zeitschrift}.

We agreed with all comments and suggestions of the referee and we have revised our submission accordingly. To simplify the second reading, we now answer the comments of the referee and summarize the changes from the last version:

{\bf Required Corrections}

\begin{itemize}
\item \textsl{\color{gray} Remark 9: I believe there are (edge) cases of pseudotriangles with only 3 points on its boundary; e.g.~take pseudotriangles in the $n = 1$ case (unless I’m supposed to count one of the boundary point twice, which isn’t clear to me).} \\
We indeed count points with multiplicity. We defined corners and hinges by the sentence ``Two consecutive wiggly arcs along the boundary of~$c$ define a \emph{corner} (resp.~a \emph{hinge}) of~$c$ if they bound a convex (resp.~concave) angle of~$c$'' in Definition 6. So two corners can indeed coincide at the same point. To insist on this fact, we have added ``Note that two corners of~$c$ can coincide at the same point~$i \in [n]$'' in Definition 6 and ``(two corners can coincide at the same point~$i \in [n]$)'' twice in Remark 9. Thank you for pointing out that this was unclear.

\item \textsl{\color{gray} Definition 34: I believe the other label should be $2h$ instead of $2h+1$.} \\
You are absolutely right, this is a typo. Thank you for noticing it.

\item \textsl{\color{gray} $g$-vectors, §3.2: This is probably the most important one. I was very confused after reading your definition of a $g$-vector; in particular, whether a $g$-vector is actually a vector in $\mathbb{H}_{2n}$. In Definition 47, you define the $g$-vector as the orthogonal projection of $\hat{g}$ onto $\mathbb{H}_{2n}$. However, the sentence right after the defining equations for $\alpha^\pm$ seems to agree with the definition that one gets before taking the projection (i.e. supposedly $\hat{g}$). I accepted this as a typo (that the $g$ should’ve been a $\hat{g}$ here), but then looking through the $g$-matrices examples that you’ve provided in the figures, I found that most of their $g$-vectors (columns) do not live in $\mathbb{H}_{2n}$ and behave more like $\hat{g}$ instead – in fact, the entries of the $g$-vectors generally did not sum to $0$, which I believe is the definition of being a vector in $\mathbb{H}_{2n}$. In reading the rest of the paper I deduced that you do want $g$-vectors to be in $\mathbb{H}_{2n}$, but please make sure that it is consistent with the examples you’ve provided.} \\
We agree with the referee that the implicit identification of the $\hat{g}$-vectors and $g$-vectors in our illustrations was really confusing. We have proceed to the following changes:
\begin{itemize}
\item We have split Figure 12 into two figures. Figure 12 now presents the $\hat{g}$-vectors and the \mbox{$g$-vectors} of a wiggly pseudotriangulation side to side, so that the projection is clearer. Figure~13 presents the $g$-vectors and the $c$-vectors of a wiggly pseudotriangulation side to side, so that the duality is visible.
\item We made clear in Example 48 and the caption of Table 1 that we represent here the $\hat{g}$-matrices. We decided to keep the $\hat{g}$-matrices rather than the $g$-matrices as the former are easier to read from Figure 5, while the latter have unnecessarily involved coefficients due to the projection. We believe that representing the $g$-matrices here would not help the reader.
\end{itemize}

@@

\item \textsl{\color{gray} pg.\,5 (C1): Perhaps more precisely, ``as arcs in D are pairwise pointed''.} \\
Indeed, this was unprecise. We modified it as suggested.

\item \textsl{\color{gray} pg.\,5 (C2)*: Possibly 3 points in the boundary, as mentioned in previous section.} \\
We have added ``(two corners can coincide at the same point~$i \in [n]$)'', see above.

\item \textsl{\color{gray} pg.\,6 (C1): As $D$ is a priori a pseudodissection, I wasn't sure what should be included in the counting of connected components. Maybe something along the lines of ``where $D$ is viewed as the set of edges of the graph with vertices given by the marked points $0$ to $n+1$'' might help.} \\
Indeed, this was too implicit. We have added ``(where $D$ is viewed as the edge set of the graph with vertex set given by the marked points $0$ to $n+1$)''.

\item \textsl{\color{gray} pg.\,6 (C2): Since $D$ is by definition a set of wiggly arcs, $c \in D$ can be confused with arcs in $D$ rather than connected components.} \\
Again, this was too implicit. The $c$ in all sums of this paragraph actually ranges over the wiggly cells of~$D$. We have added ``denoting by~$C$ the set of wiggly cells of~$D$'' and replaced~``$c \in D$'' by~``$c \in C$'' as indices of the sums.

\item \textsl{\color{gray} pg.\,6 (C3): Aren’t there examples of internal arcs which only bounds one wiggly cell? Take again $n = 1$ and the unique internal wiggly arc of any pseudotriangulation.} \\
We count with multiplicities. We have added ``(we count a wiggly arc~$\alpha$ with multiplicity~$2$ if it appears twice along the boundary of~$c$)'' to make it clear.

\item \textsl{\color{gray} pg.\,7 (C1): This is a personal tangential comment: is this a new sequence that is not currently in OEIS? If so it might be worth submitting it.} \\
Indeed, we should include it. It is not in OEIS at the moment.

\item \textsl{\color{gray} pg.\,7 (C2): I was a bit slow in understanding this; I actually had to go to the corresponding description for wiggly permutations and work my way backwards to understand this. I believe an example would help a lot.} \\
This is actually the way this decomposition was found: we first worked it out on wiggly permutations and transported it back to wiggly pseudotriangulations through the bijection.
TODO

@@

\item \textsl{\color{gray} pg.\,12 (C1): This should be ``admissible gaps''?} \\
Thanks! Fixed.

\item \textsl{\color{gray} pg.\,13 (C1): I find this description confusing. Maybe ``Following in a counterclockwise direction...'' would be better.} \\
We have replaced by ``Following each wiggly pseudotriangle of~$T$ in counterclockwise direction''.
\item \textsl{\color{gray} pg.\,13 (C2) Definition 34*: $2h + 1$ should be $2h$.} \\
You are absolutely right, we have corrected this typo.

\item \textsl{\color{gray} pg.\,14 (C1): I think it is important to first point out that every $j \in [n]$ is the hinge of exactly one pseudotriangle, which implies the highlighted sentence.} \\
Indeed, we have added ``As every~$j \in [n]$ is the hinge of precisely one pseudotriangle of~$T$''.

\item \textsl{\color{gray} Figures in §3.2 and §3.3*: As mentioned in the previous section, all $g(T)$ in figures should be replaced by the correct $g$-matrices (whose column vectors are in~$\mathbb{H}_{2n}$).} \\
We now have the $\hat{g}$-matrix and $g$-matrix side by side in Figure 12. For Table~1, we have chosen to keep the $\hat{g}$-matrices (but we made it clear in the caption). See answer above for more explanations.

### 2025-01-02 draft/wigglyhedra.tex
more corrections

\begin{proof}
Consider a wiggly pseudodissection~$D$ with~$\iota$ internal wiggly arcs, $\gamma$ wiggly cells, and~$\kappa$~connected components (where $D$ is viewed as the edge set of the graph with vertex set given by the marked points $0$ to $n+1$).
Euler's formula gives~$(n+2)-(\iota+2)+(\gamma+1)=(\kappa+1)$, hence~$n - \iota + \gamma = \kappa$.
Moreover, $\sum_{c \in D} \beta_c/2 = 1+\iota$ since each external (resp.~internal) wiggly arc bounds precisely one (resp.~two) wiggly cell of~$D$, and~${\sum_{c \in D} \big( \kappa_c-1 \big) = \kappa-1}$ since the boundary of each internal connected component of~$D$ is one of the internal connected components of the boundary of precisely one wiggly cell of~$D$ (both sums range over the wiggly cells~$c$ of~$D$).
As~$2 \le \beta_c/2+\kappa_c-1$ for any wiggly cell~$c$ by \cref{rem:degree}, we thus obtain that~$2\gamma \le \sum_{c \in D} (\beta_c/2+\kappa_c-1) = \iota+\kappa = n+\gamma$.
We conclude that~$\gamma \le n$ and~$\iota = n+\gamma-\kappa \le 2n-1$.

%Consider a wiggly pseudodissection~$D$ with~$\iota$ internal wiggly arcs, $\gamma$ wiggly cells, and~$\kappa$~connected components.
%Observe that~$\sum_{c \in D} \beta_c/2 = 1+\iota$ since each external (resp.~internal) wiggly arc bounds precisely one (resp.~two) wiggly cell of~$D$, and that~${\sum_{c \in D} \big( \kappa_c-1 \big) = \kappa-1}$ since the boundary of each internal connected component of~$D$ is one of the internal connected components of the boundary of precisely one wiggly cell of~$D$.
%As~$2 \le \beta_c/2+\kappa_c-1$ for any wiggly cell~$c$ by \cref{rem:degree}, we thus get~$2 \gamma \le \sum_{c \in D} (\beta_c/2+\kappa_c-1) = \iota+\kappa$.
%As~$D$ is a planar graph with~$v = n+2$ vertices, $e = \iota+2$ edges, $f = 1+\gamma$ faces, and $\kappa$ connected components, Euler's formula~$1+\kappa = v-e+f$ yields~$2(1+\kappa) = 2(n+2)-2(\iota+2)+2(1+\gamma) \le 2n-2\iota+2+\iota+\kappa$, so that~$\iota \le 2n-\kappa \le 2n-1$ and~$\gamma \le (\iota+\kappa)/2 \le n$.

@@

\pagebreak
\begin{remark}
Note that in the proof of \cref{prop:wigglyPseudotriangulations}, we actually obtained that
\[
\sum_{c \in D} (\delta_c-3) = \sum_{c \in D} \beta_c/2 + 2 \sum_{c \in D} (\kappa_c-1) - 2\gamma = (1+\iota) + 2(\kappa-1) - 2(\iota+\kappa-n) = 2n-1-\iota.
\]
\end{remark}

@@

\pagebreak
We will need the following observation about exchangeable arcs, illustrated in \cref{fig:incompatible3}.

@@

\enlargethispage{.3cm}
We now prove that wiggly pseudotriangulations and wiggly permutations are in bijection, and that this bijection induces an isomorphism from the wiggly flip graph to the cover graph of the wiggly lattice.
The following two definitions are illustrated in \cref{fig:wigglyLattice,fig:bijection}.

@@

\enlargethispage{.5cm}
Consider now a complete simplicial fan~$\Fan$ of~$\R^d$ with rays~$(\b{r}_v)_{v \in V}$ and cones~$\R_{\ge 0} \b{R}_\triangle$ for~${\triangle \in \Delta}$, where~$\b{R}_\triangle \eqdef \set{\b{r}_v}{v \in \triangle}$ as in \cref{prop:characterizationFan}.
To realize the fan~$\Fan$, one can try to pick a height vector~$\b{h} \eqdef (h_v)_{v \in V} \in \R^V$ and consider the polytope
\(
\polytope{P}_{\b{h}} \eqdef \set{\b{x} \in \R^d}{\dotprod{\b{r}_v}{\b{x}} \le h_v \text{ for all } v \in V}.
\)
The following classical statement characterizes the height vectors~$\b{h}$ for which the fan~$\Fan$ is the normal fan of this polytope~$\polytope{P}_{\b{h}}$.
We borrow the formulation from~\cite[Lem.~2.1]{ChapotonFominZelevinsky}.

@@

We now define two families of vectors, illustrated in \cref{fig:pseudotriangulationMatrices}.
%
\begin{figure}
\centerline{\raisebox{-1.8cm}{\includegraphics[scale=2]{pseudotriangulationMatrices}} \quad \input{figures/gcmatrices}}
\caption{The $\hat{\b{g}}$-matrix~$\hat{\b{g}}(T)$ and $4$ times the $\b{c}$-matrix~$\b{c}(T)$ of a wiggly pseudotriangulation~$T$. The $i$th column of these matrices are the $\hat{\b{g}}$-vector~$\hat{\b{g}}(\alpha)$ and $4$ times the $\b{c}$-vector~$\b{c}(\alpha, T)$ of the $i$th wiggly arc~$\alpha$ of~$T$ (ordered from bottom to top). The colors of the columns match the colors of the wiggly arcs.}
\label{fig:pseudotriangulationMatrices}
\end{figure}

@@

Observe that the $\hat{\b{g}}$-matrices of \cref{exm:specialGMatrices} have rank~$2n-1$.
More generally, we will see in \cref{coro:basis} that~$\b{g}(T)$ forms a basis of~$\HH_{2n}$ for any wiggly pseudotriangulation~$T$, which enables us to define its dual basis.

@@

\pagebreak
Recall from~\cref{rem:incompatibility-deg} that two wiggly arcs \(\alpha\) and \(\alpha'\) are exchangeable if and only if their incompatibility degree is one.
Recall from~\cref{prop:uerp} the wiggly arcs \(\beta\) and \(\beta'\) associated to \(\alpha\) and \(\alpha'\).
Finally, recall from~\cref{lem:linearDependences} that the \(\b{g}\)-vectors of the wiggly arcs \(\alpha\), \(\alpha'\), \(\beta\), and \(\beta'\) satisfy a linear dependence, depending on their type.
Let us interpret this dependence categorically.

>>>>

\begin{proof}
Consider a wiggly pseudodissection~$D$ with~$\iota$ internal wiggly arcs, $\gamma$ wiggly cells, and~$\kappa$~connected components (where $D$ is viewed as the edge set of the graph with vertex set given by the marked points $0$ to $n+1$).
Euler's formula gives~$(n+2)-(\iota+2)+(\gamma+1)=(\kappa+1)$, hence~$n - \iota + \gamma = \kappa$.
Moreover, denoting by~$C$ the set of wiggly cells of~$D$, we have~$\sum_{c \in C} \beta_c/2 = 1+\iota$ since each external (resp.~internal) wiggly arc bounds precisely one (resp.~two) wiggly cell of~$D$ (we count a wiggly arc~$\alpha$ with multiplicity~$2$ if it appears twice along the boundary of~$c$), and~${\sum_{c \in C} \big( \kappa_c-1 \big) = \kappa-1}$ since the boundary of each internal connected component of~$D$ is one of the internal connected components of the boundary of precisely one wiggly cell of~$D$.
As~$2 \le \beta_c/2+\kappa_c-1$ for any wiggly cell~$c$ by \cref{rem:degree}, we thus obtain that~$2\gamma \le \sum_{c \in C} (\beta_c/2+\kappa_c-1) = \iota+\kappa = n+\gamma$.
We conclude that~$\gamma \le n$ and~$\iota = n+\gamma-\kappa \le 2n-1$.

%Consider a wiggly pseudodissection~$D$ with~$\iota$ internal wiggly arcs, $\gamma$ wiggly cells, and~$\kappa$~connected components.
%Observe that~$\sum_{c \in C} \beta_c/2 = 1+\iota$ since each external (resp.~internal) wiggly arc bounds precisely one (resp.~two) wiggly cell of~$D$, and that~${\sum_{c \in C} \big( \kappa_c-1 \big) = \kappa-1}$ since the boundary of each internal connected component of~$D$ is one of the internal connected components of the boundary of precisely one wiggly cell of~$D$.
%As~$2 \le \beta_c/2+\kappa_c-1$ for any wiggly cell~$c$ by \cref{rem:degree}, we thus get~$2 \gamma \le \sum_{c \in C} (\beta_c/2+\kappa_c-1) = \iota+\kappa$.
%As~$D$ is a planar graph with~$v = n+2$ vertices, $e = \iota+2$ edges, $f = 1+\gamma$ faces, and $\kappa$ connected components, Euler's formula~$1+\kappa = v-e+f$ yields~$2(1+\kappa) = 2(n+2)-2(\iota+2)+2(1+\gamma) \le 2n-2\iota+2+\iota+\kappa$, so that~$\iota \le 2n-\kappa \le 2n-1$ and~$\gamma \le (\iota+\kappa)/2 \le n$.

@@

\pagebreak
\begin{remark}
Note that in the proof of \cref{prop:wigglyPseudotriangulations}, we actually obtained that
\[
\sum_{c \in C} (\delta_c-3) = \sum_{c \in C} \beta_c/2 + 2 \sum_{c \in C} (\kappa_c-1) - 2\gamma = (1+\iota) + 2(\kappa-1) - 2(\iota+\kappa-n) = 2n-1-\iota.
\]
\end{remark}

@@

We will need the following observation about exchangeable arcs, illustrated in \cref{fig:incompatible3}.

@@

We now prove that wiggly pseudotriangulations and wiggly permutations are in bijection, and that this bijection induces an isomorphism from the wiggly flip graph to the cover graph of the wiggly lattice.
The following two definitions are illustrated in \cref{fig:wigglyLattice,fig:bijection}.

@@

Consider now a complete simplicial fan~$\Fan$ of~$\R^d$ with rays~$(\b{r}_v)_{v \in V}$ and cones~$\R_{\ge 0} \b{R}_\triangle$ for~${\triangle \in \Delta}$, where~$\b{R}_\triangle \eqdef \set{\b{r}_v}{v \in \triangle}$ as in \cref{prop:characterizationFan}.
To realize the fan~$\Fan$, one can try to pick a height vector~$\b{h} \eqdef (h_v)_{v \in V} \in \R^V$ and consider the polytope
\(
\polytope{P}_{\b{h}} \eqdef \set{\b{x} \in \R^d}{\dotprod{\b{r}_v}{\b{x}} \le h_v \text{ for all } v \in V}.
\)
The following classical statement characterizes the height vectors~$\b{h}$ for which the fan~$\Fan$ is the normal fan of this polytope~$\polytope{P}_{\b{h}}$.
We borrow the formulation from~\cite[Lem.~2.1]{ChapotonFominZelevinsky}.

@@

We first define the $\b{g}$-vectors, illustrated in \cref{fig:ggMatrices}.
%
\begin{figure}
\centerline{\raisebox{-1.8cm}{\includegraphics[scale=2]{pseudotriangulationMatrices}} \quad \input{figures/ggmatrices}}
\caption{The $\hat{\b{g}}$-matrix~$\hat{\b{g}}(T)$ and $\b{g}$-matrix~$\b{g}(T)$ of a wiggly pseudotriangulation~$T$. The $i$th column of these matrices are the $\hat{\b{g}}$-vector~$\hat{\b{g}}(\alpha)$ and the $\b{g}$-vector~$\b{g}(\alpha)$ of the $i$th wiggly arc~$\alpha$ of~$T$ (ordered from bottom to top). The colors of the columns match the colors of the wiggly arcs. The $\b{g}$-vectors are the orthogonal projections of the $\hat{\b{g}}$-vectors on the hyperplane~$\HH_{2n}$.}
\label{fig:ggMatrices}
\end{figure}

@@

We will see in \cref{coro:basis} that~$\b{g}(T)$ forms a basis of~$\HH_{2n}$ for any wiggly pseudotriangulation~$T$, which enables us to define its dual basis, illustrated in \cref{fig:gcMatrices}.
%
\begin{figure}
\centerline{\raisebox{-1.8cm}{\includegraphics[scale=2]{pseudotriangulationMatrices}} \quad \input{figures/gcmatrices}}
\caption{The $\b{g}$-matrix~$\b{g}(T)$ and $\b{c}$-matrix~$\b{c}(T)$ of a wiggly pseudotriangulation~$T$. The $i$th column of these matrices are the $\b{g}$-vector~$\b{g}(\alpha)$ and $\b{c}$-vector~$\b{c}(\alpha, T)$ of the $i$th wiggly arc~$\alpha$ of~$T$ (ordered from bottom to top). The colors of the columns match the colors of the wiggly arcs. The $\b{g}$-vectors and $\b{c}$-vectors form dual bases of~$\HH_{2n}$.}
\label{fig:gcMatrices}
\end{figure}

@@

Recall from~\cref{rem:incompatibility-deg} that two wiggly arcs \(\alpha\) and \(\alpha'\) are exchangeable if and only if their incompatibility degree is one.
Recall from~\cref{prop:uerp} the wiggly arcs \(\beta\) and \(\beta'\) associated to \(\alpha\) and \(\alpha'\).
Finally, recall from~\cref{lem:linearDependences} that the \(\b{g}\)-vectors of the wiggly arcs \(\alpha\), \(\alpha'\), \(\beta\), and \(\beta'\) satisfy a linear dependence, depending on their type.
Let us interpret this dependence categorically.
